Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-11 paper/main.tex
Decent version of theory part

Experts can make mistakes. In our theoretical analysis, we will model two different types
of mistakes that the algorithm can recover from.

\begin{definition}
Let $G=(V,E_G)$ be a DAG let and $H=(V,E_H)$ be a directed graph (possibly cyclic) 
such that $E_{G^+} \subseteq E_{H^+}$. Then $\mathcal{A}_H$
is called a \emph{$G$-compatible ancestral oracle}. If $H$ is furthermore acyclic, then 
$\mathcal{A}_H$ is called a \emph{acyclic $G$-compatible ancestral oracle} with respect to $G$.
\end{definition}

These two oracles still provide correct information on all ancestral relationships that are 
present in $G$, but cannot reliably detect the \emph{absence} of ancestral relationships.
We even allow our oracles to contradict themselves by giving responses that would lead to a 
cyclic graph when combined. If we use such oracles for DAG construction, we will
introduce superfluous edges into our graph.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$, and $0$ otherwise. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \citep{Spirtes2001}.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E \cup B $ and $Y \to X \notin E \cup B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $L \gets L \cup \mathcal{A}_G(X, Y) $ \;
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    $ R \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G) $ \; 
    $ B \gets B \cup R \; ; \; L \gets L \setminus R \; ; \;  E \gets (E \cup L) \setminus B$ \;
    \Return{$(V,E,B)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycles}{FixCycles}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ X \rightarrow Y $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(X, Y, \mathbf{Z})=1$
}{
			$ R \gets R \cup \{ X \to Y \} \cup \{ Y \to X \} $ \;

@@

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a $G$-compatible ancestral oracle $\mathcal{A}_H$, 
let $\tilde{G}=\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_H, \emptyset, \infty)$
Then $H^+ \subseteq \tilde{G} \supseteq G$. If $H$ contains at least one cycle, then
$H^+ \subsetneq \tilde{G}$.
\label{prop:weakexpand:cyclic}
\end{proposition}

\begin{proof}
All edges in the skeleton of $G$ are added during $\textsc{Expand}$ with the 
correct orientation and cannot be removed by $\textsc{FixCycles}$. Therefore, the result
is a supergraph of $G$. If cycles are created during $\textsc{Expand}$, then every 
cycle contains at least one edge that is not in the skeleton of $G$, otherwise $G$ itself
would be cyclic. At least one edge of every cycle is therefore removed by
 $\textsc{FixCycles}$, making the result acyclic again. Since the removed edges need
to be in $H^+$, the first inclusion is strict for cyclic $H$.
\end{proof}

@@

\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Prune}($V$,$E$,$\mathcal{D}_G$)}{
    let $L \gets \{\}$\;
    \ForEach{$X \to Y \in E$}{
        remove $X \to Y$ from $E$\;
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 1$}{
            $L \gets L \cup  \{X \to Y\}$ \;
        }
    }
    remove all edges in $L$ from $G$\;
    \Return{$(V, E)$}\;
}
\caption{Pruning superfluous edges}
\label{algo:prune}
\end{algorithm}

@@

By combining Propositions~\ref{prop:strongexpand}, \ref{prop:weakexpand}, 
\ref{prop:weakexpand:cyclic} and 
\ref{prop:prune}, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a $G$-consistent 
ancestral oracle $\mathcal{A}_H$, 
let $(V',E',B)=\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_H,\emptyset,\infty)$.
Then $\textsc{Prune}(V',E',\mathcal{D}_G)=G.$
\end{theorem}

@@

\Fn{{\sc ExpertInLoop}($V,\mathcal{D}_G,\mathcal{A}_G$)}{
$E_p \gets \emptyset$ \tcc{Current edges} 
$B \gets \emptyset$ \tcc{Edges that were pruned or removed from cycle} 
\Repeat{ $E=E_p$ }{
	$E \gets E_p$ \;
	$(V,E,B) \gets \textsc{Expand}(V,E,\mathcal{D}_G,\mathcal{A}_G,B,1)$ \;
	$(V,E_p) \gets \textsc{Prune}(V,E,\mathcal{D}_G)$ \;
	$B \gets B \cup \{ E_p \setminus E \} $
}
\Return{(V,E)}
}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$\phi$,$B=\emptyset$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E \setminus B$ and $Y \to X \notin E \setminus B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ e_{\max} \gets \mathcal{A}_G(X, Y)$ \;
		}
        }
    }
    $L' \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G)$ \;

    \uIf{ $(V,E \cup e_{\max})$ is acyclic }{
	add $ e_{\max}$ to $E$ \;
    } \Else {
	add $ \textrm{reverse}(e_{\max})$ to $E$ \;
    }
    \Return{$(V,E)$}\;
}
\caption{Adding an edge between variables with the highest correlation}
\label{algo:ranked_expand}
\end{algorithm}

@@

Algorithm~\ref{algo:ranked_expand} shows a 
version of {\sc Expand} where we use the largest unexplained association to
determine where to next add an edge. 
To implement this, we need an effect size
metric by which we can rank effects. Depending on the type of variables, 
we can use the effect sizes shown in Section~\ref{sec:ci_tests}. 
The algorithm selects the edge that has the highest absolute residual 
association, resulting
in prioritization of edges that contribute the most to improving the model.

>>>>

Experts can make mistakes. In our analysis, we will consider experts that
essentially ``make up'' non-existing causal relationships, but do provide 
correct answers on the directionality of existing ones.

\begin{definition}
Let $G=(V,E_G)$ and $H=(V,E_H)$ be two DAGs. 
The \emph{$G$-compatible ancestral oracle} $\mathcal{A}_{G\mid H}$ is defined by
$$\mathcal{A}_{G\mid H}(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \mathcal{A}_{H}(X,Y) & \textrm{otherwise} \\
\end{cases}$$
Consider the graph $G\mid H$ containing the edges $ \mathcal{A}_{G\mid H}(X,Y)$ for all 
pairs $(X, Y) \in V \times V$.   
If $G\mid H$ is acyclic, then 
$\mathcal{A}_H$ is called an \emph{acyclic $G$-compatible ancestral oracle}.
\end{definition}

Note that a $G$-compatible oracle can contradict itself. Consider 
$V=(X,Y,Z)$, $G=(V,\{X \to Y\})$ and $H=(V,\{Y \to Z, Z \to X\})$.
This gives $\mathcal{A}_{G \mid H}(Y,Z) = Y \to Z$  
 and $\mathcal{A}_{G \mid H}(X,Z) = Z \to X$ which imply that $Y$ causes $X$, 
in contradiction to
 $\mathcal{A}_{G \mid H}(X,Y) = X \to Y$. Still, it will turn out that we can recover
from such errors.

Generally, using ancestral oracles for DAG construction will introduce superfluous edges.
Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$, and $0$ otherwise. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \citep{Spirtes2001}.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D},\mathcal{A}$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E \cup B $ and $Y \to X \notin E \cup B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}(X, Y, \mathbf{Z}) = 0$}{
	    $L \gets L \cup \mathcal{A}(X, Y) $ \;
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    $ R \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}) $ \; 
    $ B \gets B \cup R \; ; \;  E \gets (E \cup L) \setminus R$ \;
    \Return{$(V,E,B)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycles}{FixCycles}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D} $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ X \rightarrow Y $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}(X, Y, \mathbf{Z})=1$
}{
			$ R \gets R \cup \{ X \to Y \} \cup \{ Y \to X \} $ \;

@@

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a $G$-compatible ancestral oracle $\mathcal{A}_H$, 
let $\tilde{G}=\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_H, \emptyset, \infty)$
Then $G \subseteq \tilde{G} \subseteq H^+$. If $H$ contains at least one cycle, then
$\tilde{G} \subsetneq H^+$.
\label{prop:weakexpand:cyclic}
\end{proposition}

\begin{proof}
All edges in the skeleton of $G$ are added during $\textsc{Expand}$ with the 
correct orientation and cannot be removed by $\textsc{FixCycles}$. Therefore, the result
is a supergraph of $G$. If cycles are created during $\textsc{Expand}$, then every 
cycle contains at least one edge that is not in the skeleton of $G$, otherwise $G$ itself
would be cyclic. At least one edge of every cycle is therefore removed by
 $\textsc{FixCycles}$, making the result acyclic again. Since any edges removed from cycles need
to be in $H^+$, the result must be a proper subgraph of $H^+$.
\end{proof}

@@

\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Prune}($V$,$E$,$\mathcal{D}$)}{
    let $R \gets \{\}$\;
    \ForEach{$X \to Y \in E$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ 
		in $(V,E \setminus \{ X \to Y \} )$\;
        \If{$\mathcal{D}(X, Y, \mathbf{Z}) = 1$}{
            $L \gets L \cup  \{X \to Y\}$ \;
        }
    }
    remove all edges in $R$ from $G$\;
    \Return{$(V, E)$}\;
}
\caption{Pruning superfluous edges}
\label{algo:prune}
\end{algorithm}

@@

By combining Propositions~\ref{prop:strongexpand}, 
\ref{prop:weakexpand:cyclic} and 
\ref{prop:prune}, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a $G$-compatible 
ancestral oracle $\mathcal{A}_{G\mid H}$, 
let $(V',E',B)=\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_{G\mid H},\emptyset,\infty)$.
Then $\textsc{Prune}(V',E',\mathcal{D}_G)=G.$
\end{theorem}

@@

\Fn{{\sc ExpertInLoop}($V,E,\mathcal{D},\mathcal{A}$)}{
$E_p \gets \emptyset$ \tcc{Current edges} 
$B \gets \emptyset$ \tcc{Edges that were pruned or removed from cycle} 
\Repeat{ $E=E_p$ }{
	$E \gets E_p$ \;
	$(V,E,B) \gets \textsc{Expand}(V,E,\mathcal{D},\mathcal{A},B,1)$ \;
	$(V,E_p) \gets \textsc{Prune}(V,E,\mathcal{D})$ \;
	$B \gets B \cup \{ E_p \setminus E \} $
}
\Return{(V,E)}
}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D},\mathcal{A}$,$\phi$,$B$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E \cup B$ and $Y \to X \notin E \cup B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ e_{\max} \gets \mathcal{A}(X, Y)$ \;
		}
        }
    }
    $ R \gets \textsc{FixCycles}(V, E \cup \{e_{\max}\}, \mathcal{D}) $ \; 
    $ B \gets B \cup R \; ; \;  E \gets (E \cup  \{e_{\max}\}) \setminus R$ \;
    \Return{$(V,E,B)$}\;
}
\caption{Adding an edge between variables with the highest correlation}
\label{algo:rankedexpand}
\end{algorithm}

@@

Algorithm~\ref{algo:rankedexpand} shows a 
version of {\sc Expand} where we use the largest unexplained association to
determine where to next add an edge. 
To implement this, we need an effect size
metric by which we can rank effects. Depending on the type of variables, 
we can use the effect sizes shown in Section~\ref{sec:ci_tests}. 
The algorithm selects the edge that has the highest absolute residual 
association, resulting
in prioritization of edges that contribute the most to improving the model.
